---
title: "Biostrings_basics"
author: "**Authors**: Elodie Darbo et Pascal Martin"
date: "2025-04-26"
output: 
  rmdformats::material:
    
    use_bookdown: true
    thumbnails: true
    cards: false
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir="/DATA/WORK/Enseignement/2025_EBAII_ChIPseq/EBIII_N2/chip-seq/")
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(Biostrings)
```

# Introduction  

The [Biostrings](https://www.bioconductor.org/packages/release/bioc/html/Biostrings.html) package
is the core package to manipulate sequence files in R.  
Here, we will see some basic usage of this package.  

# Containers and accessors  

## XString and XStringSet containers

First, let's see how to import a [fasta](http://en.wikipedia.org/wiki/FASTA_format) file.
For this, we will use a file provided with the Biostrings package that contains
the sequences 2000 upstream of the annotated transcription start sites (TSS)
for the Drosophila melanogaster genome.  

Get the file path:
```{r}
dm3_upstream_filepath = system.file("extdata", "dm3_upstream2000.fa.gz", package="Biostrings")
```

This file contains DNA sequences so we use the `readDNAStringSet` function to import it:
```{r}
dm3_upstream = readDNAStringSet(dm3_upstream_filepath)
dm3_upstream
```

The sequences are imported as a `DNAStringSet` object.  
There are similar functions and containers to work with:  

 - RNA sequences (T becomes U): `readRNAStringSet` gives an `RNAStringSet`  
 - Protein sequences : `readAAStringSet` gives an `AAStringSet`    
 - Or any type of character strings (e.g. ASCII-encoded sequence quality): `readBStringSet` gives a `BStringSet`    
 
The `DNAStringSet` object can be considered as a list of `DNAString` objects.  
We can thus extract the 3rd sequence using:
```{r, results=3}
dm3_upstream[[3]]
# or, if we know the name of the sequence:
dm3_upstream[["NM_001201795_up_2000_chr2L_8382455_f chr2L:8382455-8384454"]]
```

Here are 2 different ways to extract a subsequence from a `DNAString`:
```{r}
dm3_upstream[[3]][2:30]
subseq(dm3_upstream[[3]], start=2, end=30)
```

But only the `subseq` command would work on a `DNAStringSet`:
```{r}
subseq(dm3_upstream[1:3], start=2, end=12)
```

It is also possible to extract multiple subsequences from a `DNAString` using `Views`:
```{r}
Views(dm3_upstream[[3]],start=c(1,11,21),end=c(10,20,30))
```
`Views` objects actually store the original sequence (or `subject`) and the intervals or `ranges` on this sequence.  
Because we were working on a `DNAString` object, the `Views` function returned an `XStringViews` object (See `?XStringViews` for details and methods).  

To convert a `DNAString` to a character string:
```{r}
toString(dm3_upstream[[3]][1:10])
# as.character also works but toString works on DNAStringSt objects too:
toString(subseq(dm3_upstream[1:2], start=1, end=10))
```


## Working with whole genome sequences: BSgenome
Whole genome sequences are stored into `BSgenome` objects.  
Many genomes are available in [Bioconductor](https://www.bioconductor.org/packages/release/BiocViews.html#___BSgenome).  
But you can also build your own using the [BSgenomeForge](https://www.bioconductor.org/packages/release/bioc/html/BSgenomeForge.html) package.

Let's load and check the dm3 BSgenome package:
```{r, message = FALSE}
library(BSgenome.Dmelanogaster.UCSC.dm3)
```
What is a bit confusing at the beginning is that loading this package creates an object named `Dmelanogaster`:
```{r}
Dmelanogaster
names(Dmelanogaster)
Dmelanogaster[["chr2L"]] #or: Dmelanogaster$chr2L
```

## Working with large fasta files: Rsamtools
The [Rsamtools](https://www.bioconductor.org/packages/release/bioc/html/Rsamtools.html) also provides interesting functions to work on large indexed FASTA files (e.g. containing a whole genome sequence).  
The short example below illustrates how to extract from a FASTA file a set of sequences defined by a `GRanges`:
```{r}
library(Rsamtools)

indFaEx_path=system.file("extdata","ce2dict1.fa",package="Rsamtools")
indFaEx=FaFile(indFaEx_path) #the whole sequence is actually not imported at this step
getSeq(indFaEx,GRanges(seqnames=Rle(c("pattern01","pattern04")),
                       ranges=IRanges(start=c(3,10),end=c(10,24)),
                       strand="*"))
```


## Some useful functions
Reverse complement:
```{r}
reverseComplement(dm3_upstream[1:2])
```

Count the occurence of each nucleotide:
```{r}
alphabetFrequency(dm3_upstream[1:2],
                  baseOnly=TRUE,
                  as.prob=TRUE)
```

Get the GC content of a sequence:
```{r}
letterFrequency(Dmelanogaster$chr2L, "CG", as.prob=TRUE)
```

We will not go into the details but there are also functions to perform sequence alignment in R ([pwalign](https://bioconductor.org/packages/release/bioc/html/pwalign.html) package):  
```{r}
pwalign::pairwiseAlignment(dm3_upstream[[1]][1:30],
                           dm3_upstream[[2]][1:30],
                           type = 'global')
```


# Searching for patterns in a sequence


# Searching for motifs in a sequence


# Exporting as fasta

